Advanced forensic Ext4 inode carving

### 2018-05-13 paper.tex
s/ext/Ext

\maketitle
\begin{abstract}
Many widely-used filesystems (NTFS, FAT, ext3) are well-researched in terms of digital forensics and data recovery, but this does not apply to ext4, a successor in the family of ext filesystems. Due to some new functionalities of ext4 and compatibility breaking, a novel approach for file carving had to developed. The advantages of this approach include its ability to restore files even in the case of a corrupted superblock. This article gives a summary of the original paper and its outcomes, also with an introduction to ext4 filesystem and ext family included.
\end{abstract}

@@

In the present, ext4 is a widely-used filesystem, especially on Linux-based operating systems and Android phones \cite{afeic}. The filesystem, released in 2008, has its origins in ext2 and ext3 filesystems. While many of their principles and structures were preserved (it is possible to upgrade from ext2 or ext3 in-place by setting new attributes), the changes are not just evolutionary. The way of storing and locating data blocks has changed, especially with the introduction of {\it extents}. From the forensical point of view, it was necessary to come up with new processes of recovering data.

This article provides a summary of the paper \cite{afeic}. But since the method is not easily understandable without wider information about ext4 way of working, we include here a description of the filesystem. Furthermore, a description of ext2 and ext3 must be also attached, as ext4 is closely tied to them and succeeds their functions.

@@

The ext4 filesystem is the next generation in the family of ext filesystems, used typically in Linux environments. In the very beginning, the Linux kernel used Minix filesystem, due to the fact that Linux itself originated from Minix. This filesystem, however, contained serious limitations, the maximum file name size and the maximum filesystem size as the most severe examples. A group of kernel developers, with Theodore Ts'o being the most prominent one, introduced a couple of new filesystems \cite{ext2design}.

The first new filesystem, ext (``extended filesystem'', released in 1992) removed the Minix limitations, but some of the previous problems remained -- for example, free blocks were monitored over a linked list, which harmed performance. As a quick response, ext2 (``second extended filesystem'', 1993) was released. While the code of ext2 originated in that of ext, it brought many notable improvements.

\subsection{The ext2 filesystem}

@@

While these features are common for all Unix filesystems, now we shall list some abilities specifical for ext2. Using the file attributes, the kernel behavior of creating new files in a directory can be modified, specifically the new file user id and group id. The logical block sizes can be specified when the filesystem is created -- larger block sizes lead to efficient behavior, but also wastes more space.

The symbolic links of ext2 can use more efficient way of storing the path inside the inode of the symlink. The overall status of the filesystem is saved in the superblock. During the mounting of the filesystem, if it is detected that it was not unmounted properly, the filesystem is automatically checked. After a certain number of mounts or when a certain time period has passed, the filesystem checking is enforced.

@@

\begin{figure}
\centering
\includegraphics[width=0.48\textwidth]{images/ext2structure.jpg}
	\caption{The structure of an ext2 filesystem \cite{ext2structure}}
\end{figure}

@@

As the main optimization, ext2 performs block readahead, that means, reads more continguous blocks at once. Block groups try to keep together inodes and blocks of a file, which reduces disk seeks. Also when a file is being written, ext2 preallocates up to 8 blocks.

\subsection{The ext3 filesystem}

After the initial period of bugs elimination, the ext2 filesystem has become a safe, stable choice for a Linux-based operating system. One of the main drawbacks of using ext2 is the lack of jounrnaling, which slows down the filesystem check and makes files prone to corruption.

First suggestions for the journaling of ext filesystem family were put down by Stephen C.\ Tweedie \cite{extjournal}. He identifies several aspects of filesystem reliability and puts forward {\it preservation} (``stable'' data on the disk should never be damaged), {\it predictability} (the failure modes of the filesystem should be predictable and deterministic) and {\it atomicity} (every operation is either fully performed or fully undone). The ext2 provided only the preservation aspect, whilst the failure states (could occur because of an unexpected reboot, for example) were not properly defined and might have prevented the operations from fully performing.

The Tweedie's article \cite{extjournal} compares approaches of reaching the desired aspects. The easiest way is to wait for a disk write to complete before the next one is submitted, which breaks the performance. Another idea lies in preserving multiple write buffers, ordering them accordingly and write these separately. However, one can easily get into a situation when the buffer dependendence is cyclic (moving a file from directory $A$ to directory $B$ and, in the same time, moving another file from $B$ to $A$). While there exists a mechanism to solve it, all the presented approaches require the disk to be completely rescanned for errors in case of a system failure. (Recall that filesystem check slowness was one of main objections against ext2.)

@@

\begin{figure}
\centering
\includegraphics[width=0.48\textwidth]{images/journal.pdf}
	\caption{The structure of ext3 journal \cite{takingadvantage}}
\end{figure}

The ext3 filesystem, released in 2001, i offers three different modes of journaling: \cite{takingadvantage}

@@

Notice that journaling cannot be turned off. Compared to ext2, this is the only different feature. Thanks to this, it is possible to convert an ext2 partition to ext3 with spawning a single command.
For the period of ext3 development, a principle of both backward and forward compatibility with ext2 was closely followed by its authors. Therefore, an ext3 partition can be mounted with ext2 driver and works the same with the exception of journaling. Turning off a journal causes an ext3 partition to become ext2 and vice versa.

\subsection{The ext4 filesystem}

Although ext3 was a step forward, it still lacked some state-of-the-art features, which could be found in other Unix filesystems. Between 2003 and 2006, numberous patches for the ext3 were proposed to be added to the Linux kernel, but they were not accepted because the Linux community was afraid of stability of their mostly-used-filesystem at that time. As a result, Theodore Ts'o introducted plans for creating a successor of ext3, which would partially break the forward compatibility, but still allowed an easy update from the older versions \cite{newext4}. In 2008, the code was marked as stable.

@@

Contrary to ext4, persistent pre-allocation of file size is possible. The checking of the filesystem is 2 to 20 times 2 to 20 times faster, as the unallocated inodes are marked and are skipped during the rest of the procedure. Instead of allocating one block at a time, the allocation is {\it delayed} in ext4, meaning it is postponed in time and performed when the file is flushed. There is also a possibility of online defragmentation. Since the size of an inode is larger, nanosecond resolution timestamps are supported and the problem of year 2038 is deferred for 272 years.

Similarly as in the case of ext2 and ext3, migrating to ext4 is just a matter of turning on the new features. The legacy system of individual block mapping is preserved, but all new files will use extents. Downgrading from ext4 to ext3 by turning off the features is straightforward when extents are not used -- otherwise, temporary copy of the files is needed.

\subsection{Comparison of ext4 and other Unix filesystems}

It is worth noting that Theodore Ts'o considered ext4 not to be a major advance and still using old technology. In the light of his words, I found potentially interesting to list other popular Unix filesystems and compare their functions.

{\it Btrfs} \cite{btrfs} is a filesystem brought to life by Oracle. The development began in 2007 and the software was proclaimed to be stable in 2014. The name comes from the B-tree as this is the main structure used within the filesystem. Notable features of Btrfs are copy-on-write (a copied and unmodified file references to the ``old data''), snapshots, CRC checksums of data, RAID or configurable compression. In order to prolong the lifespan, SSD drivers are handled specifically. In-place conversion from ext2/3/4 and ReiserFS is present.

@@

\section{File carving in ext4}

@@

Authors of the paper present an advanced method that uses patteren-based file carving . It searches for metadata structures of inodes to recover their content data. This approach avoids reading the superblock and group descriptor table since its goal is to recover files from reformatted or corrupted ext4 filesystems. The presented method can be divided into five phases\cite{dewald2017afeic}:
\begin{enumerate}
\item \textbf{Initialization}

@@

In this article, we have introduced the readers to the structure and functioning of ext2, ext3 and ext4 filesystems. For ext4, we have presented a novel approach of data recovery, which is able to restore data even in case of a corrupted superblock \cite{afeic}. The solution was proved to be successful in many situations and was adapted for Sleuthkit framework.

In future work, the authors want to add backward compatibility, so even ext4 filesystems upgraded from ext3 would be also supported. This is not possible now, as only extents are considered, not (in)direct pointers, which initially remain in a converted ext4 filesystem.

>>>>

\maketitle
\begin{abstract}
Many widely-used filesystems (NTFS, FAT, Ext3) are well-researched in terms of digital forensics and data recovery, but this does not apply to Ext4, a successor in the family of ext filesystems. Due to some new functionalities of Ext4 and compatibility breaking, a novel approach for file carving had to developed. The advantages of this approach include its ability to restore files even in the case of a corrupted superblock. This article gives a summary of the original paper and its outcomes, also with an introduction to Ext4 filesystem and ext family included.
\end{abstract}

@@

In the present, Ext4 is a widely-used filesystem, especially on Linux-based operating systems and Android phones \cite{afeic}. The filesystem, released in 2008, has its origins in Ext2 and Ext3 filesystems. While many of their principles and structures were preserved (it is possible to upgrade from Ext2 or Ext3 in-place by setting new attributes), the changes are not just evolutionary. The way of storing and locating data blocks has changed, especially with the introduction of {\it extents}. From the forensical point of view, it was necessary to come up with new processes of recovering data.

This article provides a summary of the paper \cite{afeic}. But since the method is not easily understandable without wider information about Ext4 way of working, we include here a description of the filesystem. Furthermore, a description of Ext2 and Ext3 must be also attached, as Ext4 is closely tied to them and succeeds their functions.

@@

The Ext4 filesystem is the next generation in the family of ext filesystems, used typically in Linux environments. In the very beginning, the Linux kernel used Minix filesystem, due to the fact that Linux itself originated from Minix. This filesystem, however, contained serious limitations, the maximum file name size and the maximum filesystem size as the most severe examples. A group of kernel developers, with Theodore Ts'o being the most prominent one, introduced a couple of new filesystems \cite{ext2design}.

The first new filesystem, Ext (``extended filesystem'', released in 1992) removed the Minix limitations, but some of the previous problems remained -- for example, free blocks were monitored over a linked list, which harmed performance. As a quick response, Ext2 (``second extended filesystem'', 1993) was released. While the code of Ext2 originated in that of ext, it brought many notable improvements.

\subsection{The Ext2 filesystem}

@@

While these features are common for all Unix filesystems, now we shall list some abilities specifical for Ext2. Using the file attributes, the kernel behavior of creating new files in a directory can be modified, specifically the new file user id and group id. The logical block sizes can be specified when the filesystem is created -- larger block sizes lead to efficient behavior, but also wastes more space.

The symbolic links of Ext2 can use more efficient way of storing the path inside the inode of the symlink. The overall status of the filesystem is saved in the superblock. During the mounting of the filesystem, if it is detected that it was not unmounted properly, the filesystem is automatically checked. After a certain number of mounts or when a certain time period has passed, the filesystem checking is enforced.

@@

\begin{figure}
\centering
\includegraphics[width=0.48\textwidth]{images/ext2structure.jpg}
	\caption{The structure of an Ext2 filesystem \cite{ext2structure}}
\end{figure}

@@

As the main optimization, Ext2 performs block readahead, that means, reads more continguous blocks at once. Block groups try to keep together inodes and blocks of a file, which reduces disk seeks. Also when a file is being written, Ext2 preallocates up to 8 blocks.

\subsection{The Ext3 filesystem}

After the initial period of bugs elimination, the Ext2 filesystem has become a safe, stable choice for a Linux-based operating system. One of the main drawbacks of using Ext2 is the lack of jounrnaling, which slows down the filesystem check and makes files prone to corruption.

First suggestions for the journaling of ext filesystem family were put down by Stephen C.\ Tweedie \cite{extjournal}. He identifies several aspects of filesystem reliability and puts forward {\it preservation} (``stable'' data on the disk should never be damaged), {\it predictability} (the failure modes of the filesystem should be predictable and deterministic) and {\it atomicity} (every operation is either fully performed or fully undone). The Ext2 provided only the preservation aspect, whilst the failure states (could occur because of an unexpected reboot, for example) were not properly defined and might have prevented the operations from fully performing.

The Tweedie's article \cite{extjournal} compares approaches of reaching the desired aspects. The easiest way is to wait for a disk write to complete before the next one is submitted, which breaks the performance. Another idea lies in preserving multiple write buffers, ordering them accordingly and write these separately. However, one can easily get into a situation when the buffer dependendence is cyclic (moving a file from directory $A$ to directory $B$ and, in the same time, moving another file from $B$ to $A$). While there exists a mechanism to solve it, all the presented approaches require the disk to be completely rescanned for errors in case of a system failure. (Recall that filesystem check slowness was one of main objections against Ext2.)

@@

\begin{figure}
\centering
\includegraphics[width=0.48\textwidth]{images/journal.pdf}
	\caption{The structure of Ext3 journal \cite{takingadvantage}}
\end{figure}

The Ext3 filesystem, released in 2001, i offers three different modes of journaling: \cite{takingadvantage}

@@

Notice that journaling cannot be turned off. Compared to Ext2, this is the only different feature. Thanks to this, it is possible to convert an Ext2 partition to Ext3 with spawning a single command.
For the period of Ext3 development, a principle of both backward and forward compatibility with Ext2 was closely followed by its authors. Therefore, an Ext3 partition can be mounted with Ext2 driver and works the same with the exception of journaling. Turning off a journal causes an Ext3 partition to become Ext2 and vice versa.

\subsection{The Ext4 filesystem}

Although Ext3 was a step forward, it still lacked some state-of-the-art features, which could be found in other Unix filesystems. Between 2003 and 2006, numberous patches for the Ext3 were proposed to be added to the Linux kernel, but they were not accepted because the Linux community was afraid of stability of their mostly-used-filesystem at that time. As a result, Theodore Ts'o introducted plans for creating a successor of Ext3, which would partially break the forward compatibility, but still allowed an easy update from the older versions \cite{newext4}. In 2008, the code was marked as stable.

@@

Contrary to Ext4, persistent pre-allocation of file size is possible. The checking of the filesystem is 2 to 20 times 2 to 20 times faster, as the unallocated inodes are marked and are skipped during the rest of the procedure. Instead of allocating one block at a time, the allocation is {\it delayed} in Ext4, meaning it is postponed in time and performed when the file is flushed. There is also a possibility of online defragmentation. Since the size of an inode is larger, nanosecond resolution timestamps are supported and the problem of year 2038 is deferred for 272 years.

Similarly as in the case of Ext2 and Ext3, migrating to Ext4 is just a matter of turning on the new features. The legacy system of individual block mapping is preserved, but all new files will use extents. Downgrading from Ext4 to Ext3 by turning off the features is straightforward when extents are not used -- otherwise, temporary copy of the files is needed.

\subsection{Comparison of Ext4 and other Unix filesystems}

It is worth noting that Theodore Ts'o considered Ext4 not to be a major advance and still using old technology. In the light of his words, I found potentially interesting to list other popular Unix filesystems and compare their functions.

{\it Btrfs} \cite{btrfs} is a filesystem brought to life by Oracle. The development began in 2007 and the software was proclaimed to be stable in 2014. The name comes from the B-tree as this is the main structure used within the filesystem. Notable features of Btrfs are copy-on-write (a copied and unmodified file references to the ``old data''), snapshots, CRC checksums of data, RAID or configurable compression. In order to prolong the lifespan, SSD drivers are handled specifically. In-place conversion from Ext2/3/4 and ReiserFS is present.

@@

\section{File carving in Ext4}

@@

Authors of the paper present an advanced method that uses patteren-based file carving . It searches for metadata structures of inodes to recover their content data. This approach avoids reading the superblock and group descriptor table since its goal is to recover files from reformatted or corrupted Ext4 filesystems. The presented method can be divided into five phases\cite{dewald2017afeic}:
\begin{enumerate}
\item \textbf{Initialization}

@@

In this article, we have introduced the readers to the structure and functioning of Ext2, Ext3 and Ext4 filesystems. For Ext4, we have presented a novel approach of data recovery, which is able to restore data even in case of a corrupted superblock \cite{afeic}. The solution was proved to be successful in many situations and was adapted for Sleuthkit framework.

In future work, the authors want to add backward compatibility, so even Ext4 filesystems upgraded from Ext3 would be also supported. This is not possible now, as only extents are considered, not (in)direct pointers, which initially remain in a converted Ext4 filesystem.
